\documentclass[sigplan,screen,review,natbib=true]{acmart}

%% The majority of ACM publications use class options [sigconf,review,anonymous].
%% For PLDI-adjacent workshops (like the E-Graphs Workshop), sigplan is appropriate.
%% Remove 'review' and 'anonymous' for the camera-ready version.

%% Rights management information.
\setcopyright{none}
\acmConference[EGRAPHS '26]{E-Graphs Workshop}{June 2026}{TBD}
\acmYear{2026}
\copyrightyear{2026}

%% Submission ID (remove for camera-ready).
% \acmSubmissionID{123-A56-BU3}

%% Remove ACM reference format for workshop submissions if desired.
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

%% Packages
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{stys/smtlib}

% Don't number subsubsections (acmart numbers them by default).
\setcounter{secnumdepth}{2}

\begin{document}

\title{Relational E-matching in an SMT solver}
% \subtitle{Optional Subtitle}

%% Author information. With 'anonymous' class option, these are hidden.
\author{Amar Shah}
\affiliation{%
  \institution{Carnegie Mellon University}
  \city{Pittsburgh}
  \country{USA}
}
\email{amarshah@andrew.cmu.edu}

\author{Marijn Heule}
\affiliation{%
  \institution{Carnegie Mellon University}
  \city{Pittsburgh}
  \country{USA}
}
\email{marijn@andrew.cmu.edu}

\author{Bryan Parno}
\affiliation{%
  \institution{Carnegie Mellon University}
  \city{Pittsburgh}
  \country{USA}
}
\email{bparno@andrew.cmu.edu}

\author{Max Willsey}
\affiliation{%
  \institution{University of California, Berkeley}
  \city{Berkeley}
  \country{USA}
}
\email{mwillsey@berkeley.edu}


\begin{abstract}
E-matching is the problem of finding all matches of a pattern in an egraph.
Relational e-matching, a recent approach inspired by the databases community,
has found succcess in equality saturation engines such as egglog. However, until
now, this technique has not been implemented in a Satisfiability Modulo Theories
(SMT) solver. The key difficulty of implementing relational e-matching in an SMT
solver is \emph{backtracking}, i.e., the ability to undo previous matching
decisions. We provide an initial implementation of backtrackable, relational
e-matching algorithm in Sundance, an SMT solver currently in development. We
describe our approach of maintaining backtrackable indices and our e-matching
algorithm based on a left-deep binary join. We evaluate our implementation on a
set of synthetic benchmarks and show an median speedup of $25 \times$ over a
single top-down e-matching algorithm. However, when considering a set of
benchmarks from the program verifier Verus, we see a median slowdown of $1.77
\times$. We conclude with a discussion of future work to close this gap and make
relational e-matching a practical component of an SMT solver.
\end{abstract}

%% CCS concepts and keywords (optional for workshop).
% \begin{CCSXML}
% ...
% \end{CCSXML}
% \ccsdesc[500]{Software and its engineering~Compilers}

\keywords{e-graphs, e-matching, datalog, SMT solvers}

\maketitle

\input{sections/1-intro.tex}
\input{sections/2-background}
\input{sections/3-approach}
\input{sections/4-evaluation}
\input{sections/5-related}
\input{sections/6-conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
